Genome-wide changes in expression during inflorescence architecture development

### 2015-10-14 ms.Rmd
Just plot the interesting G1-like genes

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene *GH3-8*. [**Role of auxin influx / efflux in axillary / apical meristem specification**]. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv]. Transgenic plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* (*i.e.* PBM and SBM; [@IkedaKawakatsu:2012co]), have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated to reduced auxin signalling. [**Conclusion about auxin signalling**.]

**Auxin-related genes in clusters 2 & 4:**

+ *GH3-8*, *IAA10*, *ABCB27*, *LOC_Os01g63770* (AUX/LAX transporter) in cluster 2
+ *ARF6A*, *SAUR33*, *MDR12* (ABC), *ALS1* (ABC) in cluster 4

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Only three G1-Like (G1L/ALOG) genes were detected in the dataset. Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*. [**Downstream targets of G1L5 (MADS22/MADS47/MADS55)**]

@@

```{r t_familiesPerCluster, results = "hide"}
pander::set.caption(paste0(
  I(pasteLabel("**Table ", tableCount, "t_familiesPerCluster")),
  "**: Transcription factor and transcriptional regulator family members in each cluster. *n*, total number of family members, *n*~expr~, number of family members expressed in the whole dataset, C~1~–C~", c, "~, number of family members in each cluster, and the total number of family members in all clusters."))
t_familiesPerCluster <- as.data.frame(t_familiesPerCluster[!Total == 0])
rownames(t_familiesPerCluster) <- t_familiesPerCluster$Family
t_familiesPerCluster$Family <- NULL
pander::pandoc.table(t_familiesPerCluster, split.tables = Inf, emphasize.rownames = FALSE,
                     justify = c("right", rep("centre", length(names(t_familiesPerCluster)) - 1), "left"))
```

@@

```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 2}
print(f_alogFamily)
```

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of rice G1/ALOG family genes in inflorescence meristems. Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj]. Red or blue points indicate whether the gene was detected or not detected respectively for each library. Only three G1 genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA*) were detected, but they share a similar pattern of expression.

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Neighbour-joining similarity tree of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by [@Arora:2007fa].

>>>>

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated with reduced auxin signalling. The presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. The G1-Like (G1L/ALOG) genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj]. Only three were detected in the dataset. Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*. Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55* [@Yoshida:2013ff]. *MADS22* and *MADS55* are also highly expresed in the RM, and the third rice *STMADS11*-like gene, *MADS47*, is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).

@@

@@

```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 3/2}
grid::grid.draw(f_alogFamily)
```

**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**. Expression of detected G1-Like (G1L/ALOG)  (**a**) and *STMADS11*-like MADS (**b**) genes in inflorescence meristems. Red or blue points indicate whether the gene was detected or not detected respectively for each library. Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression. Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.

@@

**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**. Neighbour-joining similarity tree of putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*. Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers. The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do]. For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes. Clades are labelled following the nomenclature presented by @Arora:2007fa.
